---
title: "Car Market Analysis in Qatar"
author: "Team 14: Aryan Arya and Tan Hai Nam Dang"
date: "2026-04-07"
format:
  revealjs:
    theme: night
    slide-number: true
    transition: fade
    background-transition: fade
    fontsize: 1.5em
---

```{r include=FALSE, messages=FALSE, warning=FALSE}
library(tidyverse)
library(here)

qatar_cars <- read_csv("../data/qatarcars.csv")
```


## Goals

- **Goal 1:** Understand how a car's country of origin influences its price in the Qatari market <br><br>

- **Goal 2:** Explore whether electric vehicles cost more than petrol cars for the same level of performance <br><br>

- **Goal 3:** Determine whether physical size and trunk volume reliably predict price, or whether brand origin and fuel type matter more

## The Dataset: Qatar Cars

- A modern, internationally-focused alternative to the classic `mtcars` dataset <br><br>

- Includes brands common in international markets, such as BMW, Mercedes, Toyota, and Honda <br><br>

- All measurements are in **International System (SI) units** (meters, kg, liters) <br><br>

- Prices are in **Qatari Riyal (QAR)**

## The Dataset: Qatar Cars

```{r echo=FALSE}
qatar_cars |>
  summarise(
    Observations = n(),
    Countries    = n_distinct(origin),
    Brands       = n_distinct(make),
    `Engine Types` = paste0(
      n_distinct(enginetype), " (Petrol, Hybrid, Electric)"
    ),
    `Price Range (QAR)` = paste0(
      scales::comma(min(price, na.rm = TRUE)), " - ",
      scales::comma(max(price, na.rm = TRUE))
    )
  ) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "Feature", values_to = "Value") |>
  knitr::kable()
```

- Key variables: `origin`, `price`, `horsepower`, `trunk`, `enginetype`, `length`, `width`, `height`, `mass`




## Car Prices by Country

- **Asian brands (China, South Korea, Japan)** cluster tightly below 200,000 QAR: affordable
- **USA** has the widest spread of any country 

```{r echo=FALSE, out.height="50%", fig.align="center"}
qatar_cars |>
  mutate(origin = fct_reorder(origin, price, median, na.rm = TRUE)) |>
  ggplot(aes(x = origin, y = price, fill = origin)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    x = "Country of Origin",
    y = "Price (QAR)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


## Car Prices by Country

- **Sweden & Germany** sit confidently in the mid-to-high range
- **Italy & UK** reach into the millions: they are status symbols

```{r echo=FALSE, out.height="50%", fig.align="center"}
qatar_cars |>
  mutate(origin = fct_reorder(origin, price, median, na.rm = TRUE)) |>
  ggplot(aes(x = origin, y = price, fill = origin)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    x = "Country of Origin",
    y = "Price (QAR, log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Origin Tells You the Price Before You See the Brand

> *In Qatar's market, knowing where a car comes from tells you almost everything about what it costs.*




## Price vs Horsepower by Engine Type

- **All three engine types start at the same place:** at low horsepower, EVs, hybrids, and petrol cars are similarly priced around 100,000–200,000 QAR
- **Petrol's slope is the steepest:** as horsepower climbs, petrol prices explode upward fastest, driven by high-performance supercars

```{r echo=FALSE, out.height="50%", fig.align="center"}
qatar_cars |>
  filter(!is.na(horsepower), !is.na(price), !is.na(enginetype)) |>
  ggplot(aes(x = horsepower, y = price, color = enginetype)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    x = "Horsepower (hp)",
    y = "Price (QAR, log scale)",
    color = "Engine Type"
  ) +
  theme_minimal()
```


## Price vs Horsepower by Engine Type

- **EVs have a flatter slope:** more horsepower does not cost as much extra compared to petrol, suggesting EVs offer better value at higher performance levels
- **Hybrids are limited in range:** few hybrid models exceed 500 hp, so their trend is less conclusive

```{r echo=FALSE, out.height="50%", fig.align="center"}
qatar_cars |>
  filter(!is.na(horsepower), !is.na(price), !is.na(enginetype)) |>
  ggplot(aes(x = horsepower, y = price, color = enginetype)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    x = "Horsepower (hp)",
    y = "Price (QAR, log scale)",
    color = "Engine Type"
  ) +
  theme_minimal()
```


## More Power Costs More, <br> But Not Equally for Every Engine

> *For the cars most people actually buy, electric and petrol cost the same. The price gap at the top is a supercar problem.*




## Vehicle Size, Price, Origin, and Engine Type: <br> PCA Analysis

- **Size variables (length, mass, width) point left:** larger, heavier cars tend to  cluster on the left, but price points **downward**, in a completely different direction
- **Price and size are not aligned:** if size drove price, their arrows would point the same way. However, they don't.

```{r echo=FALSE, out.height="50%", fig.align="center"}
pca_data <- qatar_cars |>
  select(price, length, width, height, trunk, mass, origin, enginetype) |>
  filter(complete.cases(pick(price, length, width, height, trunk, mass)))

pca_result <- pca_data |>
  select(price, length, width, height, trunk, mass) |>
  scale() |>
  prcomp()

pca_scores <- as_tibble(pca_result$x) |>
  bind_cols(pca_data |> select(origin, enginetype))

pca_loadings <- as_tibble(pca_result$rotation, rownames = "variable") |>
  mutate(PC1_scaled = PC1 * 4,
         PC2_scaled = PC2 * 4)

ggplot() +
  geom_point(data = pca_scores,
             aes(x = PC1, y = PC2, color = origin, shape = enginetype)) +
  geom_segment(data = pca_loadings,
               aes(x = 0, y = 0, xend = PC1_scaled, yend = PC2_scaled),
               arrow = arrow(length = unit(0.25, "cm"))) +
  geom_text(data = pca_loadings,
            aes(x = PC1_scaled * 1.12, y = PC2_scaled * 1.12, label = variable)) +
  labs(
    title = "PCA Biplot: Size, Price, Origin, and Engine Type",
    x = "PC1",
    y = "PC2",
    color = "Origin",
    shape = "Engine Type"
  ) +
  theme_minimal()
```


## Vehicle Size, Price, Origin, and Engine Type: <br> PCA Analysis

- **Italian & UK cars (orange & purple) sit far from the pack:** isolated at the bottom, confirming that prestige, not size, is what makes them expensive
- **Chinese & Japanese cars (green) spread rightward:** large but affordable, proving that you can have a big car without a big price tag

```{r echo=FALSE, out.height="50%", fig.align="center"}
pca_data <- qatar_cars |>
  select(price, length, width, height, trunk, mass, origin, enginetype) |>
  filter(complete.cases(pick(price, length, width, height, trunk, mass)))

pca_result <- pca_data |>
  select(price, length, width, height, trunk, mass) |>
  scale() |>
  prcomp()

pca_scores <- as_tibble(pca_result$x) |>
  bind_cols(pca_data |> select(origin, enginetype))

pca_loadings <- as_tibble(pca_result$rotation, rownames = "variable") |>
  mutate(PC1_scaled = PC1 * 4,
         PC2_scaled = PC2 * 4)

ggplot() +
  geom_point(data = pca_scores,
             aes(x = PC1, y = PC2, color = origin, shape = enginetype)) +
  geom_segment(data = pca_loadings,
               aes(x = 0, y = 0, xend = PC1_scaled, yend = PC2_scaled),
               arrow = arrow(length = unit(0.25, "cm"))) +
  geom_text(data = pca_loadings,
            aes(x = PC1_scaled * 1.12, y = PC2_scaled * 1.12, label = variable)) +
  labs(
    title = "PCA Biplot: Size, Price, Origin, and Engine Type",
    x = "PC1",
    y = "PC2",
    color = "Origin",
    shape = "Engine Type"
  ) +
  theme_minimal()
```

## Size Does Not Sets the Price, But Origin Does

> *A bigger car is not a more expensive car. In Qatar's market, brand prestige matters more than size.*




## The Brand Name Is The Most Powerful Pricing Factor

- **Goal 1 ✓** <br> Country of origin is the strongest price signal in Qatar's market. Asian brands dominate the affordable segment, European brands own the premium tier, while Italian & UK cars exist in a league of their own. <br><br>

- **Goal 2 ✓** <br> EVs are not more expensive than petrol for equivalent performance. At usual horsepower levels, the price gap is negligible. Therefore, the EV revolution is more affordable than you could imagine. <br><br>

- **Goal 3 ✓** <br> Size does not reliably predict price. A large Chinese SUV and a compact Italian supercar can differ by millions. Prestige sets the ceiling, not dimensions. 

